The goal of {sitrep} is provide report templates for common epidemiological
surveys and outbreak reports. The package further contains helper function that
standardize certain analyses. 

While templates are primarily for Medecins Sans Frontieres analyses - they have 
been setup to be as generic as possible for use by the general applied 
epidemiology community. 

**Detailed information about the project, the templates and functions in {sitrep} 
can be found at <https://r4epi.github.io/sitrep/>.**

{sitrep} includes a number of other *R* packages which facilitate specific analysis:  
[{epitabulate}](https://r4epi.github.io/epitabulate/): Tables for epidemiological analysis  
[{epidict}](https://r4epi.github.io/epidict/): Epidemiology data dictionaries and random data generators  
[{epikit}](https://r4epi.github.io/epikit/): Miscellaneous helper tools for epidemiologists  
[{apyramid}](https://r4epi.github.io/apyramid/): Age pyramid construction and plotting  


## Installation

You can install {sitrep} from CRAN:

``` r
install.packages("sitrep")
```

<details>
<!--
NOTE: everything inside the details tag will be collapsed and effectively
hidden from the user
-->
<summary style='text-decoration: underline'>Click here for alternative installation options</summary>

You can also install the in-development version from GitHub using the {remotes}
package (but there's no guarantee that it will be stable):

```{r eval = FALSE}
# install.packages("remotes")
remotes::install_github("R4EPI/sitrep") 
```

</details>
```
 
## Available templates

Sitrep has outbreak templates and survey templates available. These templates 
will generate the following:

1. A word document with the situation report
2. A plain text markdown document (for conversion to other formats such as HTML or PDF)
3. A directory with all of the figures produced

You can access the list of templates in *R Studio* by clicking (see example below): 
file > New file > R Markdown... > From Template

You can generate an example template by using the `check_sitrep_templates()`
function:

```{r check_templates, warning = FALSE, message = FALSE}
library("sitrep")
output_dir <- file.path(tempdir(), "sitrep_example")
dir.create(output_dir)

# view the available templates, categorized by type
available_sitrep_templates(categorise = TRUE)

# generate the measles outbreak template in the output directory
check_sitrep_templates("measles_outbreak", path = output_dir)

# view the contents
list.files(output_dir, recursive = TRUE)
```


